Beyond clines: lineages and haplotype blocks in hybrid zones

### 2016-04-22 drafts/review-responses-april-2016.tex

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We did all the things, thanks.

\plr{these aren't points because they are all points below in Reviewer 1 \& 2's sections; we'll just say here that we did 'em, see below}

@@

\begin{quote}
Reviewer 1 would like to see development of the finding that ancestral blocks are longer around a selected locus into a practical test.  I think that this would be a substantial project for the future, and that it is valuable to first explain and understand the theoretical prediction, as is done here.  Those analysing actual data can of course look for this pattern without needing an already implemented test.
\end{quote}

\begin{quote}
Rev. 1 is also concerned that only under dominance is simulated.  However, it is clear that any form of selection that gives the same marginal selection coefficient will give essentially the same effect: a large body of work on clines shows that different selective mechanisms are indistinguishable, and it is reasonable for the authors to rely on that.  The other main concern was about the novelty of Eqs 4, 5; these are indeed standard (which I think the authors intended by referring to them as the Kolmogorov backward equations).  I don't know of a reference where they have been written in exactly this form, but the recombination terms and the diffusion terms in Eq. 4 both go back a long way.
\end{quote}

\begin{quote}
Reviewer 2 is mainly concerned about the fit between theory and simulations.  I am not worried about this, since the deterministic mathematics stands independently, and the numerical computations are straightforward.  The stochastic simulations are useful mainly as illustrations of the process.  Theory and simulations are compared in the supplementary material; it should be stated more clearly how closely these match. (A precise fit is not expected anyway because there is substantial drift).  Rev. 2 also worries that the simulations are over a rather narrow range; I suspect that the authors should have results that address sensitivity to the boundaries.  Regarding specific point 1, I don't know of a previous derivation of exactly these equations, but similar equations have been used widely, and I don't think that he authors suggest otherwise.
\end{quote}

@@

\reply
\alisa{Our decision to keep the derivations in appendices was with consideration for the readership of Molecular Ecology}
\plr{good start...}

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
    This manuscript presents a discussion of patterns of linkage and allele frequency clines in a hybrid zone between two populations, under a scenario of mild hybrid incompatibility. It presents deterministic equations for the distribution of ancestral haplotypes along the genome and argues that the lengths of ancestry segments can be used to infer selection.

    The idea of using the lengths of ancestry blocks to detect selection in hybrid zones is interesting, and I had not encountered it before. This article will likely be of most interest to theorists, however, as the discussion of the practical confounders of the selection test is limited. 
\end{quote}

\begin{point}{}
    A serious limitation of the test is that it only detects selection against heterozygotes. The claim that this would not affect results, on l.453-456:
    \textit{``Although our assumed scenario of selection against heterozygotes at a single locus is uncommon in nature, we believe that this is unlikely to drastically influence our findings, as previous studies have demonstrated that underdominant loci share similar properties with more realistic models''}
    is unconvincing.
\end{point}

\reply
\alisa{We agree with the comments of the AE on the matter, and have added some additional text to the introduction that hopefully convinces that this is ok.}

\begin{point}{}
    From a theoretical perspective, the article makes modest but valid contributions, and  Equation 4 and 5 are potentially interesting extensions to equation (1). The practical benefit of using these equations rather than forward simulations appears limited.
\end{point}

\reply
\alisa{This is a point we have considered ourselves. We present these equations as a way of building our model up to address haplotype lengths. Both analytical expressions and simulations are presented so that they each support the conclusions of the other.}

\begin{point}{}
    While reading the manuscript, I found it difficult to distinguish the most important observations from more tangential observations. The manuscript reads like a long list of observations, and would benefit from more focus on the main findings and more structure.
\end{point}

@@

\reply

\begin{point}{}
Since the recombination-based terms in Eqs 4 and 5 do not depend on the spatially explicit part of the model or on the selection term, I would be surprised if they had not been derived somewhere.
\end{point}

\reply
\plr{yaniv doing that}

\begin{point}{Eq.5}
If $a<b<\theta<0$, then I interpret $g(\theta,b) = 0$. If correct, this should be specified.  Alternately, the bounds of the integral could be modified to account for this situation.
\end{point}

\reply
\plr{peter did this}

@@

\begin{point}{}
As an example of a section that felt unstructured, consider the discussion starting at l.347.  This long paragraph starts with a long list of observations, without giving a hint as to where we are going. Beginning the paragraph with the main idea for the paragraph would be useful.
\end{point}

\reply
\alisa{Working on how to fix this in the ms now}

\begin{point}{``In the notation above, $q_z(x, t, r) = P^x{Z_t = A}$.''}
I did not see the notation $P^x$ above, and the RHS does not depend on small z.
\end{point}

\reply
\alisa{(This is at l180) Should the $=$ be a $:=$}
\plr{peter will fix this}

\begin{point}{} 
``As we show below the solutions provide an good approximation''
\end{point}

@@

% Quality of Science: Experimentally and/or theoretically excellent reliable data, no fatal flaws
% 
% Importance of Science: Research addresses a consequential question in ecology, evolution, behaviour, or conservation
% 
% Quality of Presentation: Ideas and methods mostly clear, but grammar and/or spelling is poor, format does not follow guidelines, and /or there is redundancy between sections
% 
% Does this manuscript require significant reduction in length? If 'Yes', please indicate where shortening is required in your specific comments: Yes
% 
% Does the Data Accessibility section list all the datasets needed to recreate the results in the manuscript? If `No', please specify which additional data are needed in your comments to the authors.: Yes

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
    In this paper, the authors extend previous work on the diffusion of neutral alleles in a secondary-contact zone to include selection against hybrids. Building on the classic diffusion model for the dynamics of a cline at a selected locus, they formalize the dynamics of linked neutral loci and investigate the distribution of ancestry block-lengths around the selected locus.

    The authors obtain numerical results for the flattening of the cline at a neutral locus linked to an under-dominant selected locus in the absence of genetic drift. They also outline scaling arguments for the rates of movement and length of the introgressing blocks around the selected locus. Simulations are set up to test the theoretical predictions and illustrate other interesting patterns such as the distribution of haplotype (ancestry block?) lengths.
\end{quote}

\begin{point}{}
    Unfortunately, the fit of the theoretical predictions to simulations is largely not shown, and those comparisons in the manuscript are limited to the frequency dynamics of the neutral alleles (Figure S3). The predictive theoretical results are based on scaling arguments - but it is not clear which of these results are uniquely derived in this paper.
\end{point}

\reply
\alisa{Our hope was that the simulations and theory would provide qualitative support for one another. Some predictions made by the theory e.g. time and spatial scales do seem to correspond. We will try to do a better job of signposting our unique contributions??}

\plr{peter to add sentence saying this is standard, ref slatkin, barton}

\plr{alisa to see if we can add any more from what we've done}

\begin{point}{}
    The authors also aim to characterize the movement of the genomic block that contains a selected locus through a hybrid zone. Yet, the size of the simulated habitat barely covers the cline width, and the simulation results may be sensitive to the boundaries.

Overall, this MS presents some interesting insights, and the formalization of the block dynamics can be useful. However, I think that more work is needed to explain and test (by simulations) the novel predictions. Currently, the theory and the matching simulations are largely presented separately.
\end{point}

\reply
\alisa{Should we be running more sims? Nick seems to think we have done this, but I don't think we have.}
\plr{run sims with longer range}
\plr{also run sims in part of param range where nothing's supposed to happen?}
\plr{alisa to add $s=.001$ and the like}

@@

\begin{point}{Figure 1}
 Plotting the haplotypes to the side of the plot would help.
\end{point}

\reply
\alisa{We have remade Figure 1, with additional information that we hope will help with interpretation}
\plr{alisa to refer to these events in discussion of terms in figure 5}

@@

\begin{point}{l. 245}
 As the theory is set up in the absence of genetic drift, and, correspondingly, the local population size is large, it is unclear how a deme with 500 individuals could become empty in a Wright-Fisher model.  The (narrow) width of the habitat also needs justification.
\end{point}

\reply
\alisa{It is true that the demes are never empty. However, this is the way the simulation is coded in order to account for the possibility if smaller population sizes are ever simulated.}
\plr{peter to add note about width of sims in the section describing simulations}

@@

\begin{point}{Figure 3:}
What are the replicate lines for the same $r$? Why re the lines for $r = 0$ all slightly displaced to the right? Also, it would appear that the color scheme in associated Figure S2 (bottom) is mismatched with the legend, shifted to the red part of the spectrum. After fixing this, I think that the Figure S2 should be in the main text instead of Fig. 3.
\end{point}

\reply
\alisa{We believe that the displacement is likely due to drift. (We could do a repeated sim to show this?)}

\plr{alisa to clarify figure legend}
\plr{peter or alisa fix S2}

@@

\begin{point}{l. 299}
Total population size is irrelevant when predicting the effects of genetic drift (besides, some simulations only have 50 demes across, not 100). In a one-dimensional habitat, one uses $N \sigma \sqrt{s}$ to judge whether the effect of genetic drift is likely to be negligible.
\end{point}

\reply
\alisa{We are concerned here with linked loci, so the $N\sigma\sqrt{s}$ rule of thumb does not necessarily apply here? Also, within demes, population size is much smaller (500?)}
\plr{I agree: they are probably thinking of selected loci, not linked ones. We're talking about global drift not local drift. Yaniv to investigate.}

@@

\reply
\plr{see above; add something to discussion.}

\alisa{We should add a line or two in the discussion about drift. However, IBD is perhaps not relevant over the timescales we are considering and is not necessarily ancestry informative, and since we are not interested in the equilibrium state, it is not clear that we need to worry about this limit. }

\begin{point}{p.27 rates, e.g. Eq. (7,8):}
 I find the explanation of the rates a bit unclear. I understand that the recombinant AB can arise either by A reproducing at rate $s_A$ and recombining with B, or B reproducing at rate $s_B$ and recombining with A -- such that the neutral locus of the type A is kept. Or alternatively, the BA recombinant arises, where the neutral locus at B is kept. Which one of the two happens has a probability of 1/2.
\end{point}

\reply
\alisa{We'll work on making this more understandable.}
\plr{peter to do this}

\begin{point}{}
The ``Data Accessibility section'' is missing. The folder at \href{http://github.com_petrelharp_clinal-lineages}  appears to be a working folder with lots of other files and the r-scripts hidden under a `sim' directory. The README file appears insufficient to allow orientation in the scripts, and worryingly, it has an `oh, wrong recombination distance' note to it.
\end{point}

\reply
\alisa{Ok. This probably needs to be fixed.}
\plr{alisa to write a good README: what scripts to run to get a simulation; where each figure gets produced}

\begin{point}{}
The Acknowledgements section is missing.
\end{point}

### 2016-04-23 drafts/hybrid_zone.tex
peter went through responses once

There is substantial theoretical work on genetic clines.
We focus on those maintained only by selection against hybrids -- \emph{tension zones} --
as in \citet{Bazykin1969}, reviewed in \citet{Barton1985}.
Clines may also be maintained by environmental gradients \citep{Haldane1948,fisher1950frequencies};
although these are in principle different,
\citet{Slatkin1973} pointed out, in a far-reaching general analysis,
that while the environment may determine the location of a cline,
the fitness of the heterozygote will likely determine the shape of the cline,
and \citet{barton1986barrier} show that \alisa{clines under several different forms of selection are similar in shape.}
There is substantial work on how clines are affected by genetic drift \citep{Slatkin1975,felsenstein1975genetic,barton2008effect,Polechova2011},
and quantifying the barrier to gene flow caused by clines \citep{barton1979gene,Barton1989,Barton1993,Barton2000} \alisa{and much of this focuses on equilibrium states}.
In this work, we focus on \emph{transient} patterns after the initial formation of a tension zone,
in particular describing haplotype patterns.
Although the selected cline establishes quickly, transient patterns can persist for quite some time,
and \alisa{ancestry} haplotypes can potentially carry substantial information about the formation of recent zones.
To do this, we take the reverse-time perspective,
understanding the temporal dynamics of the system by describing properties of the \emph{lineages} tracing back from the zone.
This is particularly suited to understanding transient patterns 
although, because of the well-known problems with spatial coalescence \citep{felsenstein1975pain,barton2002continuous},
incorporating genetic drift is not straightforward.
Common to nearly all studies is the use of the
reaction-diffusion equations that govern the deterministic, high-density limit \citep[as in][]{Nagylaki1975},
or the discrete analogues \citep{hanson1966effects}.

@@

If hybrid ancestry at a given locus is disfavored, migrant haplotypes containing the selected locus will be removed rapidly from the population, preventing introgression of surrounding genomic regions. We therefore expect a deficit of short blocks of foreign ancestry surrounding the selected locus, with this effect becoming more pronounced further away from the center of a hybrid zone. As a corollary to this, conditional on having the ancestry that is at lower frequency (that is, being on the `wrong' side of the hybrid zone), the length of unbroken ancestry surrounding the selected locus is expected to be relatively long when far away from the zone center.  This is because an unfit haplotype is more likely to have been recently inherited from the other side of the hybrid zone, and therefore will not have
as many ancestors of the locally common type as do neutral haplotypes. \alisa{This reasoning is applicable to both instances of under dominance and ecological (e.g. local adaptation) }

@@

In this section we aim to give both heuristics for how lineages move, 
that are helpful in providing intuition and for establishing order-of-magnitude estimates,
and analytics, mostly in terms of partial differential equations needing numerical solution. \alisa{For simplicity, we ignore coalescence.}

\paragraph{Establishment of the cline at the selected locus}
Deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline,
affecting neighboring loci as well \citep{barton1979gene, Barton1986}.
First, we turn our attention to how the cline at the selected locus itself is formed. 
\alisa{This has been previously described, but is presented here as it provides a framework for further development of our model}
Write $p(x,t)$ for the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and the $A$ allele is initially fixed on the left, so
$p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
We assume that dispersal is local and unbiased;
and that the mean squared displacement between parent and offspring
in the direction perpendicular to the cline
is $\sigma^2$.
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous,
and (iii) population density is large,
% and (iv) $s$ and $\sigma$ are small, \alisa{relative to $N$ and $\tau$ respectively?}
then as in \citet{Bazykin1969},
the commonly-used equation that $p$ approximately solves is
\begin{align} \label{eqn:cline_pde}
    \frac{d}{dt} p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\Delta p$ is the Laplacian.
\plr{change $\Delta$ to the second derivative?}
The equation is approximate because it 
omits terms of order $s^2$
(so deviations from the prediction occur over $1/s^2$ generations),
and relies on the Central Limit Theorem to approximate generation-to-generation dispersal by the Gaussian
(so if dispersal is non-Gaussian, will fit best over longer time scales, say, tens of generations).  \revpoint{AE}{2}
This says that the local rate of change of the frequency of $A$ ($\frac{d}{dt} p$)
is equal to the net rate at which $A$ alleles arrive by local dispersal ($\sigma^2/2 \Delta p$)
and a logistic term proportional to the difference in reproductive rate between $A$ and $a$ alleles
due to local proportions of heterozygotes,
which is $s(2p-1)$ (up to an error of size $s^2$).
This can be derived as in \citet{Nagylaki1975};
for discussion of diploidy see \citet{christiansen1995genotypic}.
As we show below the solutions provide a good approximation across the range of realistic values of $s$.
Although there is no known exact analytic solution to this equation, 
the steady-state solution is
$\lim_{t \to \infty} p(x,t) = (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. 
This relies on several approximations, but the general conclusions should be quite robust: 
the stable cline has width of order $\sigma/\sqrt{s}$, and decays exponentially. 
Rescaling space and time by $\sigma/\sqrt{s}$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} 
results in a dimensionless equation \citep{Slatkin1973}, \linelabel{ll:rescaling}
implying that the cline establishes over a timescale of the order $1/s$. 
Since this is how long it takes diffusion at rate $\sigma^2$ to smooth across a region of width $\sigma/\sqrt{s}$,
this means we can think of
the cline as established by diffusion, 
despite being slowed down somewhat by selection against heterozygotes.

@@

Consider the collection of $A$ alleles found at geographic location $x$.
The expected number, among these alleles, that had a parent at location $y$ in the previous generation 
is proportional to the number of $A$ alleles that were at $y$ multiplied by the per-generation probability of dispersing from $y$ to $x$. 
In other words, lineages move as a random walk determined by the dispersal kernel,
but biased towards locations where the selected allele they carry is at higher frequency.
(Lineages are also biased towards regions with higher fecundity,
and thus away from the hybrid zone, but we assume that $s$ is small and ignore this.) \alisa{could this reference to small $s$ be confusing?}

Making the same assumptions as for equation \eqref{eqn:cline_pde}, the derivation in Appendix \ref{apx:lineage_derivation} shows 
that the lineage of an $A$ allele moves as Brownian motion driven by dispersal
with mean displacement proportional to the derivative of $\log p$.
This has the effect of pulling $A$ lineages towards regions where they are more common
% with speed $\sigma$
% with bias $\sigma^2 \grad \log p(x,t)$, where $\grad$ is the gradient
(i.e., it moves at speed $\sigma^2$ in the potential $-\log p$). \alisa{What does potential here mean?}
% \alisa{note to self to read through appendix when written}
To see why this is true, note that the probability that the parent of an $A$ allele found at $x$
lived at position $x-r$ is proportional to $\P\{R=-r\} p(x-r)$, where $R$ is the random dispersal distance;
and so the mean displacement from offspring to parent is $-\E[R p(x-R)]/\E[p(x-R)]$.
% $$ = \E[ -R ] - \E[R]^2 p'(x)/p(x) + \E[R^2] p'(x)/p(x) + O(R^3) $$
Using the fact that $\E[R]=0$ and $\E[R^2]=\sigma^2$ \alisa{perhaps a brief explanation of how this works}
and expanding $p()$ to first order about $x$ shows that this mean displacement is approximately $\sigma^2 p'(x)/p(x)$,
which is $\sigma^2$ multiplied by the gradient of $\log p(x)$.
This description holds even when the frequency profile of the $A$ allele changes with time (replacing $p(x)$ by $p(x,t)$).

@@

We can describe this process formally in It\^o notation:
with $B_\tau$ a standard Brownian motion,
$T_B(\tau)$ the most recent time before $\tau$ that $Z_\tau = B$, and likewise for $T_A$,
\begin{align}
    \begin{aligned} \label{eqn:lineage_motion}
        dX_\tau &= dB_\tau + \begin{cases}
             \sigma^2 \grad \log(p(X_\tau,\tau)) d\tau \qquad & \text{if } Z_\tau = A \\
             \sigma^2 \grad \log(1-p(X_\tau,\tau)) d\tau \qquad & \text{if } Z_\tau = B 
        \end{cases} \\
        % \P\{ Z_{\tau+\epsilon} = B \given Z_\tau = A \} &= \epsilon r (1-p(X_\tau,\tau)) + O(\epsilon^2) \\
        % \P\{ Z_{\tau+\epsilon} = A \given Z_\tau = B \} &= \epsilon r p(X_\tau,\tau) + O(\epsilon^2)  .
        \P\{ T_B(\tau) &> \tau+u \given Z_\tau = A \} = \exp\left( - r \int_\tau^{\tau+u} (1-p(X_s,s)) ds \right) \\
        \P\{ T_A(\tau) &> \tau+u \given Z_\tau = B \} = \exp\left( - r \int_\tau^{\tau+u} p(X_s,s) ds \right) .
    \end{aligned}
\end{align}
\alisa{Explain in words}
In the first expression, giving the distribution of how the lineage location changes,
the first term ($\sigma dB_\tau$) is Brownian noise driven by dispersal,
and the second is the mean displacement, which moves the lineage ``downhill'' on either $-\log p$ or $-\log (1-p)$
($\grad$ is the gradient),
depending on which selected allele the lineage is linked to.
The second and third expressions say that a lineage switches from an $A$ to $B$ background
at rate $r$ multiplied by the local proportion of $B$ alleles, which is $(1-p)$.

@@

\paragraph{Linked clines}
We can use this diffusion model for lineages to find expected clines in ancestry,
i.e., the expected proportion of individuals who inherit from species $A$,
as a function of space, time, and position on the genome.
Precisely, we need the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$,
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
We denote this probability $q_z(x,t,r)$.
In the notation above,
\begin{align}
    q_z(x,t,r) = \P^x \{Z_t = A\} .
\end{align}
(Recall that $Z$ depends implicitly on $r$.)
The description above implies that $q_z$ solves the following Kolmogorov backward equation:
\begin{align}
    \begin{aligned}  \label{eqn:q_pde}
    \frac{d}{dt} q_A(x,t,r) 
            &= \sigma^2 \grad \log(p(x,t)) \cdot \grad q_A(x,t,r) 
                + \frac{\sigma^2}{2} \Delta q_A(x,t,r) 
            \\ &\qquad {} + 
                r (1-p(x,t))(q_B(x,t,r)-q_A(x,t,r))  \\
    \frac{d}{dt} q_B(x,t,r) &= \sigma^2 \grad \log(1-p(x,t)) \cdot \grad q_B(x,t,r) 
            + \frac{\sigma^2}{2} \Delta q_B(x,t,r)
            \\ &\qquad {} + 
            r p(x,t) (q_A(x,t,r)-q_B(x,t,r))  ,
    \end{aligned} 
\end{align}
\alisa{Use our words}
with boundary conditions $q_A(x,0,r)=1$ and $q_B(x,0,r)=0$, 
and where $\grad$ is the gradient, and $\Delta$ is the Laplacian \alisa{describing the rate of change}. 
The three terms in these equations come from Brownian movement of a linages,
the tendency of a lineage to inherit from regions where its type is more common,
and recombination between selected backgrounds, respectively.
Equivalently,
the first term ($\Delta q$) reflects the smoothing action of local dispersal,
the second is the net flux induced by reduced fitness in the hybrid zone,
and the third is the contribution of recombination.
%\alisa{this is fwd-in-time notation. should we change it to bwd-in-time?}
%\plr{This equation should be forwards in time, since it's the Kolmogorov backward equation
%applied to a backwards-time process!}

@@

\subsection*{Single-loci: clines}

>>>>

There is substantial theoretical work on genetic clines.
We focus on those maintained only by selection against hybrids -- \emph{tension zones} --
as in \citet{Bazykin1969}, reviewed in \citet{Barton1985}.
Clines may also be maintained by environmental gradients \citep{Haldane1948,fisher1950frequencies};
although these are in principle different,
\citet{Slatkin1973} pointed out, in a far-reaching general analysis,
that while the environment may determine the location of a cline,
the fitness of the heterozygote will likely determine the shape of the cline,
and \citet{barton1986barrier} show that \alisa{clines under several different forms of selection are similar in shape.}
\plr{(What forms of selection?)}
There is substantial work on how clines are affected by genetic drift \citep{Slatkin1975,felsenstein1975genetic,barton2008effect,Polechova2011},
and quantifying the barrier to gene flow caused by clines \citep{barton1979gene,Barton1989,Barton1993,Barton2000} \alisa{and much of this focuses on equilibrium states}.

In this work, we focus on \emph{transient} patterns after the initial formation of a tension zone,
in particular describing haplotype patterns.
Although the selected cline establishes quickly, transient patterns can persist for quite some time,
and \alisa{ancestry} haplotypes can potentially carry substantial information about the formation of recent zones.
\plr{(I think we just want to say ``haplotypes'', because ``ancestry haplotypes'' isn't a thing, and although we do only study the sort of information about haplotype patterns we can get by painting by ancestry, the statement holds in general.)}
To do this, we take the reverse-time perspective,
understanding the temporal dynamics of the system by describing properties of the \emph{lineages} tracing back from the zone.
This is particularly suited to understanding transient patterns 
although, because of the well-known problems with spatial coalescence \citep{felsenstein1975pain,barton2002continuous},
incorporating genetic drift is not straightforward.
Common to nearly all studies is the use of the
reaction-diffusion equations that govern the deterministic, high-density limit \citep[as in][]{Nagylaki1975},
or the discrete analogues \citep{hanson1966effects}.

@@

If hybrid ancestry at a given locus is disfavored, migrant haplotypes containing the selected locus will be removed rapidly from the population, preventing introgression of surrounding genomic regions. We therefore expect a deficit of short blocks of foreign ancestry surrounding the selected locus, with this effect becoming more pronounced further away from the center of a hybrid zone. As a corollary to this, conditional on having the ancestry that is at lower frequency (that is, being on the `wrong' side of the hybrid zone), the length of unbroken ancestry surrounding the selected locus is expected to be relatively long when far away from the zone center.  This is because an unfit haplotype is more likely to have been recently inherited from the other side of the hybrid zone, and therefore will not have
as many ancestors of the locally common type as do neutral haplotypes. \alisa{This reasoning is applicable to both instances of under dominance and ecological (e.g. local adaptation).}
\plr{(ecological what?)}

@@

In this section we aim to give both heuristics for how lineages move, 
that are helpful in providing intuition and for establishing order-of-magnitude estimates,
and analytics, mostly in terms of partial differential equations needing numerical solution. 
\alisa{For simplicity, we ignore coalescence.}
\plr{(say why and what that means: e.g., deterministic dynamics)}

\paragraph{Establishment of the cline at the selected locus}
Deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline,
affecting neighboring loci as well \citep{barton1979gene, Barton1986}.
First, we turn our attention to how the cline at the selected locus itself is formed. 
\alisa{This has been previously described, but is presented here as it provides a framework for further development of our model.}
\plr{(I presume it's been described, but I don't think we found a reference that actually did this, right?  So maybe say it's ``well-known in the field'' or something?  Also, no-one seems to have looked at lineages, and as our final point says, we actually get some more information out of this perspective.)}
Write $p(x,t)$ for the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and the $A$ allele is initially fixed on the left, so
$p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
We assume that dispersal is local and unbiased;
and that the mean squared displacement between parent and offspring
in the direction perpendicular to the cline
is $\sigma^2$.
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous,
and (iii) population density is large,
% and (iv) $s$ and $\sigma$ are small, \alisa{relative to $N$ and $\tau$ respectively?}
then as in \citet{Bazykin1969},
the commonly-used equation that $p$ approximately solves is
\begin{align} \label{eqn:cline_pde}
    \frac{d}{dt} p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\Delta p$ is the Laplacian.
\plr{change $\Delta$ to the second derivative?}
The equation is approximate because it 
omits terms of order $s^2$
(so deviations from the prediction occur over $1/s^2$ generations),
and relies on the Central Limit Theorem to approximate generation-to-generation dispersal by the Gaussian
(so if dispersal is non-Gaussian, will fit best over longer time scales, say, tens of generations).  \revpoint{AE}{2}
This says that the local rate of change of the frequency of $A$ ($\frac{d}{dt} p$)
is equal to the net rate at which $A$ alleles arrive by local dispersal ($\sigma^2/2 \Delta p$)
and a logistic term proportional to the difference in reproductive rate between $A$ and $a$ alleles
due to local proportions of heterozygotes,
which is $s(2p-1)$ (up to an error of size $s^2$).
This can be derived as in \citet{Nagylaki1975};
for discussion of diploidy see \citet{christiansen1995genotypic}.
As we show below the solutions provide a good approximation across the range of realistic values of $s$.
Although there is no known exact analytic solution to this equation, 
the steady-state solution is
$\lim_{t \to \infty} p(x,t) = (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. 
This relies on several approximations, but the general conclusions should be quite robust: 
the stable cline has width of order $\sigma/\sqrt{s}$, and decays exponentially. 
Rescaling space and time by $\sigma/\sqrt{s}$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} 
results in a dimensionless equation,
implying that the cline establishes over a timescale of the order $1/s$. 
(This sort of rescaling is common in the literature, e.g., \citep{fisher1950frequencies,Slatkin1973}.)  \linelabel{ll:rescaling}
Since this is how long it takes diffusion at rate $\sigma^2$ to smooth across a region of width $\sigma/\sqrt{s}$,
this means we can think of
the cline as established by diffusion, 
despite being slowed down somewhat by selection against heterozygotes.

@@

Consider the collection of $A$ alleles found at geographic location $x$.
The expected number, among these alleles, that had a parent at location $y$ in the previous generation 
is proportional to the number of $A$ alleles that were at $y$ multiplied by the per-generation probability of dispersing from $y$ to $x$. 
In other words, lineages move as a random walk determined by the dispersal kernel,
but biased towards locations where the selected allele they carry is at higher frequency.
% (Lineages are also biased towards regions with higher fecundity,
% and thus away from the hybrid zone, but we assume that $s$ is small and ignore this.) 
% \alisa{could this reference to small $s$ be confusing?}
% \plr{it's confusing because it's differences in fecundity that bias the lineage since they create the net flux of A alleles to the right, even though there's a different term in how a lineage moves for the difference in previous-generation fecundity, that we're ignoring. I'm removing the statement.}

Making the same assumptions as for equation \eqref{eqn:cline_pde}, the derivation in Appendix \ref{apx:lineage_derivation} shows 
that the lineage of an $A$ allele moves as Brownian motion driven by dispersal
with mean displacement proportional to the derivative of $\log p$.
This has the effect of pulling $A$ lineages towards regions where they are more common
% with speed $\sigma$
% with bias $\sigma^2 \grad \log p(x,t)$, where $\grad$ is the gradient
(the analogous physical system is a particle that moves at speed $\sigma^2$ in the potential $-\log p$). 
\alisa{What does potential here mean?}
\plr{(It means like a gravity well.  Clear now that it says ``particle''?)}
To see why this is true, note that the probability that the parent of an $A$ allele found at $x$
lived at position $x-r$ is proportional to $\P\{R=-r\} p(x-r)$, where $R$ is the random dispersal distance;
and so the mean displacement from offspring to parent is $-\E[R p(x-R)]/\E[p(x-R)]$.  \revpoint{1}{9}
% $$ = \E[ -R ] - \E[R]^2 p'(x)/p(x) + \E[R^2] p'(x)/p(x) + O(R^3) $$
Using the fact that $\E[R]=0$ and $\E[R^2]=\sigma^2$ 
\alisa{(perhaps a brief explanation of how this works)}
\plr{(That's what the rest of the sentence is? Work it out, and if you still think it needs more detail, add it?)}
and expanding $p()$ to first order about $x$ shows that this mean displacement is approximately $\sigma^2 p'(x)/p(x)$,
which is $\sigma^2$ multiplied by the gradient of $\log p(x)$.
This description holds even when the frequency profile of the $A$ allele changes with time (replacing $p(x)$ by $p(x,t)$).

@@

We can describe this process formally in It\^o notation:
with $B_\tau$ a standard Brownian motion,
$T_B(\tau)$ the most recent time before $\tau$ that $Z_\tau = B$, and likewise for $T_A$,
\begin{align}
    \begin{aligned} \label{eqn:lineage_motion}
        dX_\tau &= dB_\tau + \begin{cases}
             \sigma^2 \grad \log(p(X_\tau,\tau)) d\tau \qquad & \text{if } Z_\tau = A \\
             \sigma^2 \grad \log(1-p(X_\tau,\tau)) d\tau \qquad & \text{if } Z_\tau = B 
        \end{cases} \\
        % \P\{ Z_{\tau+\epsilon} = B \given Z_\tau = A \} &= \epsilon r (1-p(X_\tau,\tau)) + O(\epsilon^2) \\
        % \P\{ Z_{\tau+\epsilon} = A \given Z_\tau = B \} &= \epsilon r p(X_\tau,\tau) + O(\epsilon^2)  .
        \P\{ T_B(\tau) &> \tau+u \given Z_\tau = A \} = \exp\left( - r \int_\tau^{\tau+u} (1-p(X_s,s)) ds \right) \\
        \P\{ T_A(\tau) &> \tau+u \given Z_\tau = B \} = \exp\left( - r \int_\tau^{\tau+u} p(X_s,s) ds \right) .
    \end{aligned}
\end{align}
\alisa{Explain in words.}
\plr{(I hope that comment came before I added these words? Or do you want different ones?)}
In the first expression, giving the distribution of how the lineage location changes,
the first term ($\sigma dB_\tau$) is Brownian noise driven by dispersal,
and the second is the mean displacement, which moves the lineage ``downhill'' on either $-\log p$ or $-\log (1-p)$
($\grad$ is the gradient),
depending on which selected allele the lineage is linked to.
The second and third expressions say that a lineage switches from an $A$ to $B$ background
at rate $r$ multiplied by the local proportion of $B$ alleles, which is $(1-p)$.

@@

\paragraph{Linked clines}
We can use this diffusion model for lineages to find expected clines in ancestry,
i.e., the expected proportion of individuals who inherit from species $A$,
as a function of space, time, and position on the genome.
Precisely, we need the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$,
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
We denote this probability $q_z(x,t,r)$.
In the notation above,
\begin{align}
    q_z(x,t,r) = \P^x \{Z_t = A \given X_0 = x, \; Z_0 = z\} .
\end{align}
(Recall that $Z$ depends implicitly on $r$.) \revpoint{1}{7}
The description above implies that $q_z$ solves the following Kolmogorov backward equation:
\begin{align}
    \begin{aligned}  \label{eqn:q_pde}
    \frac{d}{dt} q_A(x,t,r) 
            &= \sigma^2 \grad \log(p(x,t)) \cdot \grad q_A(x,t,r) 
                + \frac{\sigma^2}{2} \Delta q_A(x,t,r) 
            \\ &\qquad {} + 
                r (1-p(x,t))(q_B(x,t,r)-q_A(x,t,r))  \\
    \frac{d}{dt} q_B(x,t,r) &= \sigma^2 \grad \log(1-p(x,t)) \cdot \grad q_B(x,t,r) 
            + \frac{\sigma^2}{2} \Delta q_B(x,t,r)
            \\ &\qquad {} + 
            r p(x,t) (q_A(x,t,r)-q_B(x,t,r))  ,
    \end{aligned} 
\end{align}
\alisa{Use our words} \plr{(what does that mean)}
with boundary conditions $q_A(x,0,r)=1$ and $q_B(x,0,r)=0$, 
and where $\grad$ is the gradient, and $\Delta$ is the Laplacian 
\alisa{describing the rate of change}. 
\plr{(the Laplacian is describing the rate of change? I don't follow.)}
The three terms in these equations come from Brownian movement of a linages,
the tendency of a lineage to inherit from regions where its type is more common,
and recombination between selected backgrounds, respectively.
Equivalently,
the first term ($\Delta q$) reflects the smoothing action of local dispersal,
the second is the net flux induced by reduced fitness in the hybrid zone,
and the third is the contribution of recombination.
%\alisa{this is fwd-in-time notation. should we change it to bwd-in-time?}
%\plr{This equation should be forwards in time, since it's the Kolmogorov backward equation
%applied to a backwards-time process!}

@@

\subsection*{Single locus clines}
